---
title: "AML SYK inhibitor response analysis"
author: "Ehsan Zangene"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    df_print: paged
runtime: shiny
params:
  excel_file: "../data/MS Figure data_4.7.2026.xlsx"
  output_dir: "results"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = F,
  warning = FALSE,
  message = FALSE,
  fig.width = 9,
  fig.height = 5,
  dpi = 150
)
```

# 1. Packages

```{r packages}
required_pkgs <- c("rlang",
  "readxl", "writexl", "tidyverse", "janitor", "stringr", "glue",
  "broom", "broom.mixed", "pracma", "lme4", "lmerTest", "emmeans",
  "patchwork", "scales", "DT", "plotly", "viridis", "pheatmap"
)

installed <- rownames(installed.packages())
for (pkg in setdiff(required_pkgs, installed)) {
  install.packages(pkg)
}

invisible(lapply(required_pkgs, library, character.only = TRUE))

params = list(
  excel_file = "../data/MS Figure data_4.7.2026.xlsx",
  output_dir = "results" )

dir.create(params$output_dir, showWarnings = FALSE, recursive = TRUE)
```

# 2. Project paths

```{r paths}
excel_path <- params$excel_file
stopifnot(file.exists(excel_path))

fig_dir  <- file.path(params$output_dir, "figures")
tab_dir  <- file.path(params$output_dir, "tables")
rds_dir  <- file.path(params$output_dir, "rds")

dir.create(fig_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(tab_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(rds_dir, showWarnings = FALSE, recursive = TRUE)
```

# 3. Read every sheet from the Excel workbook

```{r read-all-sheets}
excel_sheets_vec <- readxl::excel_sheets(excel_path)

raw_sheets <- setNames(
  lapply(excel_sheets_vec, function(sh) readxl::read_excel(excel_path, sheet = sh)),
  excel_sheets_vec
)

clean_names_df <- function(df) {
  df %>% janitor::clean_names()
}

clean_sheets <- lapply(raw_sheets, clean_names_df)

tibble(clean_sheets$Fig6_PA)

saveRDS(raw_sheets, file.path(rds_dir, "raw_sheets.rds"))
saveRDS(clean_sheets, file.path(rds_dir, "clean_sheets.rds"))

sheet_inventory <- tibble(
  sheet = names(clean_sheets),
  nrow  = map_int(clean_sheets, nrow),
  ncol  = map_int(clean_sheets, ncol),
  columns = map_chr(clean_sheets, ~ paste(names(.x), collapse = ", "))
)

write_xlsx(list(sheet_inventory = sheet_inventory), file.path(tab_dir, "sheet_inventory.xlsx"))
sheet_inventory

fig6_pa = clean_sheets$Fig6_PA
```

```{r clean-data}
stopifnot(!is.null(fig6_pa))

fig6_pa <- fig6_pa %>%
  as_tibble() %>%
  clean_names() %>%
  mutate(
    flow_time = as.numeric(flow_time),
    treat = factor(treat, levels = c("UT", "ENTO", "LANRA")),
    npm1_genot = factor(npm1_genot, levels = c("WT", "mut")),
    rel_p_syk = as.numeric(rel_p_syk),
    dose_n_m = readr::parse_number(dosage),
    dose_unit = stringr::str_extract(dosage, "[A-Za-z]+$"),
    dose_unit = if_else(is.na(dose_unit), "nM", dose_unit),
    dose_nM = case_when(
      dose_n_m == 0 ~ 0,
      dose_unit == "nM" ~ dose_n_m,
      dose_unit == "uM" ~ dose_n_m * 1000,
      dose_unit == "mM" ~ dose_n_m * 1e6,
      TRUE ~ dose_n_m
    ),
    dose_label = factor(dosage, levels = unique(dosage))
  )

fig6_pa
```

# Structure check

```{r structure-check}
fig6_pa %>% count(flow_time, treat, npm1_genot, dosage)
```

# Reconstruct sample IDs

There is **no explicit sample ID column** in this sheet. To calculate per-sample dose-response curves and AUC, I need one. The code below reconstructs a `sample_id` under the following assumption:

-   within each `flow_time × npm1_genot × treat` block,
-   observations are ordered by sample,
-   and each sample contributes one value per dose.

For example, for `ENTO` at 2 hour flow read in `WT` genotype, the first 3 doses are assumed to belong to sample 1 and the next 3 doses to sample 2.

```{r reconstruct-sample-id}
# Number of non-zero doses within each time/treat block
n_doses_lookup <- fig6_pa %>%
  filter(treat != "UT") %>%
  distinct(flow_time, treat, dosage, dose_nM) %>%
  count(flow_time, treat, name = "n_doses")

fig6_with_id <- fig6_pa %>%
  left_join(n_doses_lookup, by = c("flow_time", "treat")) %>%
  group_by(flow_time, npm1_genot, treat) %>%
  mutate(row_in_group = row_number()) %>%
  ungroup() %>%
  mutate(
    sample_id = case_when(
      treat == "UT" ~ row_in_group,
      TRUE ~ ceiling(row_in_group / n_doses)
    ),
    sample_key = paste(flow_time, npm1_genot, sample_id, sep = "_")
  ) %>%
  select(-row_in_group)

fig6_with_id %>% arrange(flow_time, npm1_genot, treat, sample_id, dose_nM)
```

# QC check for inferred sample IDs

```{r qc-sample-id}
fig6_with_id %>%
  count(flow_time, npm1_genot, treat, sample_id) %>%
  arrange(flow_time, npm1_genot, treat, sample_id)
```

# Convert to inhibition metric

`rel_p_syk` is already normalized to untreated. A lower value means stronger inhibition. For intuition, we also define:

-   `inhibition = 1 - rel_p_syk`

So the larger inhibition the stronger drug response.

```{r add-inhibition}
fig6_with_id <- fig6_with_id %>%
  mutate(
    inhibition = 1 - rel_p_syk,
    log2_rel_p_syk = log2(rel_p_syk)
  )

fig6_with_id
```

# Quick descriptive summary

```{r descriptive-summary}
summary_tbl <- fig6_with_id %>%
  filter(treat != "UT") %>%
  group_by(flow_time, treat, npm1_genot, dosage, dose_nM) %>%
  summarise(
    n = n(),
    median_rel_p_syk = median(rel_p_syk, na.rm = TRUE),
    mean_rel_p_syk = mean(rel_p_syk, na.rm = TRUE),
    sd_rel_p_syk = sd(rel_p_syk, na.rm = TRUE),
    median_inhibition = median(inhibition, na.rm = TRUE),
    .groups = "drop"
  )

summary_tbl
write.csv(summary_tbl, file.path("results", "tables", "fig6_pa_summary_by_condition.csv"), row.names = FALSE)
```

# Plot 1: individual dose-response trajectories

This is the clearest replacement for the current bar plot. It shows:

-   each inferred sample trajectory,
-   all individual points,
-   and genotype-level median trend lines.

```{r plot-dose-response, fig.width=11, fig.height=6}
# install.packages("ggh4x") # if needed
library(ggh4x)

p_dose <- fig6_with_id %>%
  filter(treat != "UT") %>%
  ggplot(aes(x = dose_nM, y = rel_p_syk, color = npm1_genot)) +
  geom_hline(yintercept = 1, linetype = "dotted", linewidth = 0.5) +
  
  # Sample lines
  geom_line(
    aes(group = interaction(sample_key, treat), linetype = "Sample"),
    alpha = 0.35, linewidth = 0.7
  ) +
  
  # Sample points
  geom_point(aes(shape = "Sample"), size = 2.4) +
  
  # Median line
  stat_summary(
    aes(group = npm1_genot, linetype = "Median"),
    fun = median,
    geom = "line",
    linewidth = 1.2
  ) +
  
  # Median point
  stat_summary(
    aes(group = npm1_genot, shape = "Median"),
    fun = median,
    geom = "point",
    size = 3.2
  ) +
  
  ggh4x::facet_grid2(
    rows = vars(flow_time),
    cols = vars(treat),
    scales = "free_x",
    independent = "x",
    labeller = labeller(
      flow_time = function(x) paste0(x, " h"),
      treat = label_value
    )
  ) +
  
  ggh4x::facetted_pos_scales(
    x = list(
      flow_time == 2 ~ scale_x_continuous(
        trans = pseudo_log_trans(base = 10),
        breaks = c(1, 10, 30),
        labels = c("1", "10", "30"),
        expand = expansion(mult = c(0.08, 0.25))
      ),
      flow_time == 24 ~ scale_x_continuous(
        trans = pseudo_log_trans(base = 10),
        breaks = c(200, 1000, 5000),
        labels = c("200", "1000", "5000"),
        expand = expansion(mult = c(0.08, 0.25))
      )
    )
  ) +
  
  geom_text(
    aes(
      label = sprintf("%.2f", rel_p_syk),
      color = after_scale(scales::alpha(colour, 1))
    ),
    position = position_jitter(width = 0.1, height = 0.3),
    hjust = -0.15,
    vjust = 0.3,
    size = 3,
    show.legend = FALSE
  ) +
  scale_shape_manual(
    name = "Point type",
    values = c("Sample" = 16, "Median" = 18)
  ) +
  
  labs(
    title = "Figure 6A: dose-response of relative pSYK by NPM1 genotype",
    x = "Dose (uM)",
    y = "Relative pSYK (treated / untreated)",
    color = "NPM1"
  )

p_dose

ggsave(file.path("./results/figures/", "fig6_pa_dose_response11_6_.png"), p_dose, width = 11, height = 6, dpi = 300)
```

# Plot 2: same data as inhibition

Some readers find this easier to interpret because higher values mean stronger response.

```{r plot-inhibition, fig.width=11, fig.height=6}
# install.packages("ggh4x") # if needed
library(ggh4x)

p_inhib <- fig6_with_id %>%
  filter(treat != "UT") %>%
  ggplot(aes(x = dose_nM, y = inhibition, color = npm1_genot)) +
  geom_hline(yintercept = 0, linetype = "dotted", linewidth = 0.5) +
  
  # Sample lines
  geom_line(
    aes(group = interaction(sample_key, treat), linetype = "Sample"),
    alpha = 0.35, linewidth = 0.7
  ) +
  
  # Sample points
  geom_point(aes(shape = "Sample"), size = 2.4) +
  
  geom_text(
    aes(label = sprintf("%.2f", rel_p_syk)),
    hjust = -0.15,
    vjust = 0.3,
    size = 3,
    show.legend = FALSE
  ) +
  
  # Median line
  stat_summary(
    aes(group = npm1_genot, linetype = "Median"),
    fun = median,
    geom = "line",
    linewidth = 1.2
  ) +
  
  # Median point
  stat_summary(
    aes(group = npm1_genot, shape = "Median"),
    fun = median,
    geom = "point",
    size = 3.2
  ) +
  
  ggh4x::facet_grid2(
    rows = vars(flow_time),
    cols = vars(treat),
    scales = "free_x",
    independent = "x",
    labeller = labeller(
      flow_time = function(x) paste0(x, " h"),
      treat = label_value
    )
  ) +
  
  ggh4x::facetted_pos_scales(
    x = list(
      flow_time == 2 ~ scale_x_continuous(
        trans = pseudo_log_trans(base = 10),
        breaks = c(1, 10, 30),
        labels = c("1", "10", "30"),
        expand = expansion(mult = c(0.08, 0.25))
      ),
      flow_time == 24 ~ scale_x_continuous(
        trans = pseudo_log_trans(base = 10),
        breaks = c(200, 1000, 5000),
        labels = c("200", "1000", "5000"),
        expand = expansion(mult = c(0.08, 0.25))
      )
    )
  ) +
  
  # Legend definitions
  scale_linetype_manual(
    name = "Line type",
    values = c("Sample" = "solid", "Median" = "solid")
  ) +
  scale_shape_manual(
    name = "Point type",
    values = c("Sample" = 16, "Median" = 18)
  ) +
  
  labs(
    title = "Figure 6A: inhibition view of pSYK response",
    x = "Dose (uM)",
    y = "Inhibition = 1 - relative pSYK",
    color = "NPM1"
  )

p_inhib

ggsave(file.path("results", "figures", "fig6_pa_inhibition_response.png"), p_inhib, width = 11, height = 6, dpi = 300)
```

# Publication-oriented compact figure

This figure combines the main dose-response view and the AUC summary.

```{r combined-figure, fig.width=12, fig.height=9}
# p_compact <- p_dose / p_auc + plot_layout(heights = c(2, 1))
# p_compact

p_compact_2 <- p_dose / p_inhib

p_compact_2

ggsave(file.path("results", "figures", "fig6_pa_compact_publication_style.png"), p_compact, width = 12, height = 9, dpi = 300)
```



# Suggested interpretation notes

```{r interpretation-notes, results='asis'}
cat("\n")
cat("- A lower relative pSYK value indicates stronger SYK inhibition.\n")

```


