---
title: "Figure 6A dose-response, v2"
author: "Ehsan Zangene"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8.8,
  fig.height = 6.2,
  dpi = 300
)
```

# Purpose

This document rebuilds the old `results/figures/fig6_pa_dose_response11_6_.png`
as a cleaner Figure 6A panel from the `Fig 6` sheet in
`data/MS Figure data_5.4.2026.xlsx`.

The attached Tania conversation and PDF mockup point toward a figure that:

- keeps the same 2 h and 24 h flow dose-response story,
- uses ENTO and LANRA as the two drug columns,
- emphasizes NPM1 mutant response with a strong teal line,
- keeps WT visible in coral,
- shows individual sample trajectories faintly and the genotype median boldly,
- labels the plotted values directly so the reader does not need a table.

```{r packages}
required_packages <- c(
  "readxl", "dplyr", "tidyr", "ggplot2", "patchwork", "scales",
  "stringr", "readr", "ggrepel", "ggh4x"
)

missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages) > 0) {
  stop(
    "Please install the missing package(s) first: ",
    paste(missing_packages, collapse = ", "),
    "\nExample: install.packages(c(",
    paste(sprintf("\"%s\"", missing_packages), collapse = ", "),
    "))",
    call. = FALSE
  )
}

library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
library(scales)
library(stringr)
library(readr)
library(ggrepel)
library(ggh4x)
```

```{r paths}
input_file <- file.path("data", "MS Figure data_5.4.2026.xlsx")
sheet_name <- "Fig 6"
output_dir <- file.path("results", "figures")
table_dir <- file.path("results", "tables")

if (!file.exists(input_file) && file.exists("MS Figure data_5.4.2026.xlsx")) {
  input_file <- "MS Figure data_5.4.2026.xlsx"
}

if (!file.exists(input_file)) {
  stop(
    "Input file not found: ", input_file,
    "\nPut 'MS Figure data_5.4.2026.xlsx' inside the project's data/ folder.",
    call. = FALSE
  )
}

dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(table_dir, recursive = TRUE, showWarnings = FALSE)
```

# Read Figure 6A Blocks

The `Fig 6` sheet is formatted as a manuscript figure worksheet rather than a
tidy table. The Figure 6A data are in two blocks:

- rows 7-13: total-cell flow relative pSYK at 2 h,
- rows 20-26: total-cell flow relative pSYK at 24 h.

Columns 4-5 are NPM1 WT samples, and columns 7-8 are NPM1 mutant samples.

```{r parse-figure-6a}
raw_fig6 <- readxl::read_excel(input_file, sheet = sheet_name, col_names = FALSE)

parse_fig6a_block <- function(raw, row_idx, flow_time) {
  block <- raw[row_idx, , drop = FALSE]

  tibble(
    treat = as.character(block[[1]]),
    dosage = as.character(block[[3]]),
    wt_1 = suppressWarnings(as.numeric(block[[4]])),
    wt_2 = suppressWarnings(as.numeric(block[[5]])),
    mut_1 = suppressWarnings(as.numeric(block[[7]])),
    mut_2 = suppressWarnings(as.numeric(block[[8]]))
  ) |>
    filter(!is.na(treat), treat %in% c("ENTO", "LANRA")) |>
    pivot_longer(
      cols = c(wt_1, wt_2, mut_1, mut_2),
      names_to = "sample_code",
      values_to = "rel_p_syk"
    ) |>
    filter(!is.na(rel_p_syk)) |>
    mutate(
      flow_time = flow_time,
      npm1_genot = if_else(str_detect(sample_code, "^wt"), "WT", "mut"),
      sample_id = case_when(
        sample_code == "wt_1" ~ "WT sample 1",
        sample_code == "wt_2" ~ "WT sample 2",
        sample_code == "mut_1" ~ "mut sample 1",
        sample_code == "mut_2" ~ "mut sample 2",
        TRUE ~ sample_code
      ),
      dose_nM = readr::parse_number(dosage),
      dose_uM = dose_nM / 1000,
      dose_axis = dose_nM,
      dose_label = case_when(
        flow_time == 2 ~ as.character(dose_nM),
        TRUE ~ as.character(dose_nM)
      ),
      condition_id = paste(flow_time, treat, npm1_genot, sample_id, sep = "_"),
      inhibition = 1 - rel_p_syk
    )
}

fig6_pa_long <- bind_rows(
  parse_fig6a_block(raw_fig6, 7:13, 2),
  parse_fig6a_block(raw_fig6, 20:26, 24)
) |>
  mutate(
    treat = factor(treat, levels = c("ENTO", "LANRA")),
    npm1_genot = factor(npm1_genot, levels = c("WT", "mut")),
    flow_time = factor(flow_time, levels = c(2, 24), labels = c("2 h", "24 h"))
  ) |>
  arrange(flow_time, treat, npm1_genot, sample_id, dose_axis)

fig6_pa_summary <- fig6_pa_long |>
  group_by(flow_time, treat, npm1_genot, dose_axis, dose_label) |>
  summarise(
    n = n(),
    median_rel_p_syk = median(rel_p_syk, na.rm = TRUE),
    mean_rel_p_syk = mean(rel_p_syk, na.rm = TRUE),
    sd_rel_p_syk = sd(rel_p_syk, na.rm = TRUE),
    median_inhibition = median(inhibition, na.rm = TRUE),
    .groups = "drop"
  )

readr::write_csv(fig6_pa_long, file.path(table_dir, "fig6_pa_panel_a_long_data.csv"))
readr::write_csv(fig6_pa_summary, file.path(table_dir, "fig6_pa_panel_a_summary_v2.csv"))

fig6_pa_long
```

# Figure Style

```{r figure-style}
npm1_colours <- c("WT" = "#F06D67", "mut" = "#05B9C2")
npm1_light <- c("WT" = alpha("#F06D67", 0.28), "mut" = alpha("#05B9C2", 0.28))

theme_fig6a <- function(base_size = 10.5) {
  theme_classic(base_size = base_size) +
    theme(
      axis.line = element_line(color = "grey20", linewidth = 0.45),
      axis.ticks = element_line(color = "grey20", linewidth = 0.35),
      axis.text = element_text(color = "grey20"),
      axis.title = element_text(face = "bold", color = "grey10"),
      panel.grid.major.y = element_line(color = "grey88", linewidth = 0.35),
      panel.grid.minor = element_blank(),
      strip.placement = "outside",
      strip.text.x = element_text(face = "bold", size = 11, color = "grey12"),
      strip.text.y = element_text(face = "bold", size = 11, color = "grey12"),
      legend.position = "right",
      legend.title = element_text(face = "bold"),
      legend.key = element_rect(fill = "white", color = NA),
      plot.title = element_text(face = "bold", size = 12, hjust = 0),
      plot.margin = margin(6, 8, 6, 8),
      plot.background = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA)
    )
}
```

# Panel A: Dose Response

```{r figure-6a-dose-response, fig.width=8.8, fig.height=6.2}
label_df <- fig6_pa_long |>
  mutate(
    label = sprintf("%.2f", rel_p_syk),
    label_nudge_x = case_when(
      dose_axis == min(dose_axis) ~ -0.02,
      TRUE ~ 0.06
    )
  )

p_fig6a <- ggplot(
  fig6_pa_long,
  aes(x = dose_axis, y = rel_p_syk, color = npm1_genot)
) +
  geom_hline(yintercept = 1, linetype = "dashed", linewidth = 0.35, color = "grey25") +
  geom_line(
    aes(group = condition_id, color = npm1_genot),
    linewidth = 0.75,
    alpha = 0.30
  ) +
  geom_point(
    aes(group = condition_id, color = npm1_genot),
    size = 2.25,
    alpha = 0.95
  ) +
  stat_summary(
    aes(group = npm1_genot, color = npm1_genot),
    fun = median,
    geom = "line",
    linewidth = 1.65
  ) +
  stat_summary(
    aes(group = npm1_genot, color = npm1_genot),
    fun = median,
    geom = "point",
    shape = 18,
    size = 3.3
  ) +
  ggrepel::geom_text_repel(
    data = label_df,
    aes(label = label, color = npm1_genot),
    size = 3.0,
    fontface = "bold",
    min.segment.length = Inf,
    box.padding = 0.10,
    point.padding = 0.05,
    max.overlaps = Inf,
    show.legend = FALSE,
    seed = 612
  ) +
  ggh4x::facet_grid2(
    rows = vars(flow_time),
    cols = vars(treat),
    scales = "free_x",
    independent = "x",
    strip = ggh4x::strip_themed(
      background_x = ggh4x::elem_list_rect(
        fill = c("#9EB9EA", "#F4B6B5"),
        color = NA
      ),
      background_y = ggh4x::elem_list_rect(
        fill = c("#D7D7D7", "#D7D7D7"),
        color = NA
      )
    )
  ) +
  ggh4x::facetted_pos_scales(
    x = list(
      flow_time == "2 h" ~ scale_x_continuous(
        trans = scales::pseudo_log_trans(base = 10),
        breaks = c(1, 10, 30),
        labels = c("1", "10", "30"),
        expand = expansion(mult = c(0.22, 0.20))
      ),
      flow_time == "24 h" ~ scale_x_continuous(
        trans = scales::pseudo_log_trans(base = 10),
        breaks = c(200, 1000, 5000),
        labels = c("200", "1000", "5000"),
        expand = expansion(mult = c(0.22, 0.20))
      )
    )
  ) +
  scale_y_continuous(
    limits = c(0, 1.12),
    breaks = c(0, 0.25, 0.50, 0.75, 1.00),
    expand = expansion(mult = c(0.03, 0.08))
  ) +
  scale_color_manual(
    name = "NPM1",
    values = npm1_colours,
    labels = c("WT", "mut")
  ) +
  guides(
    color = guide_legend(
      override.aes = list(linewidth = 1.6, size = 3.2, alpha = 1)
    )
  ) +
  labs(
    title = "Figure 6A: dose-response of relative pSYK by NPM1 genotype",
    x = "Dose (nM)",
    y = "Relative pSYK (treated / untreated)"
  ) +
  theme_fig6a()

p_fig6a

ggsave(
  file.path(output_dir, "fig6_pa_dose_response_v2.png"),
  p_fig6a,
  width = 8.8,
  height = 6.2,
  dpi = 450,
  bg = "white"
)

ggsave(
  file.path(output_dir, "fig6_pa_dose_response_v2.pdf"),
  p_fig6a,
  width = 8.8,
  height = 6.2,
  bg = "white"
)
```

# Optional Layout Proof

The attached PDF mockup shows duplicate "Panel A" and "Panel B" blocks. This
extra proof stacks two copies of the dose-response panel so the layout can be
tested in a larger figure assembly without changing the underlying plot.

```{r optional-stacked-layout-proof, fig.width=8.8, fig.height=12.8}
p_panel_a <- p_fig6a + plot_annotation(tag_levels = "A")
p_panel_b <- p_fig6a + plot_annotation(tag_levels = "B")

p_stacked_proof <- p_panel_a / p_panel_b +
  plot_layout(heights = c(1, 1), guides = "collect") &
  theme(legend.position = "right")

p_stacked_proof

ggsave(
  file.path(output_dir, "fig6_pa_dose_response_v2_stacked_layout_proof.png"),
  p_stacked_proof,
  width = 8.8,
  height = 12.8,
  dpi = 350,
  bg = "white"
)
```

# Notes For Manuscript Interpretation

```{r interpretation, results='asis'}
cat("- Lower relative pSYK means stronger pSYK inhibition after treatment.\n")
cat("- The visual emphasis is on whether NPM1 mutant samples drop more than WT samples across dose.\n")
cat("- Individual samples are retained as faint lines; the bold diamond trajectory is the genotype median.\n")
```
